---
title: "Windowing"
author: "Brian J. Knaus"
date: "`r Sys.Date()`"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = 'center')
knitr::opts_chunk$set(fig.width = 8)
knitr::opts_chunk$set(fig.height = 6)
```


The package vcfR was designed to work on individual chromosomes.
This is because a chromosome is a convenient way to subset a genome into smaller tasks.
We are frequently interested in examining an entire genome in order to identify features of interest.
This can be accomplished by processing chromosomes one at a time and writing summaries to a file.
This allows us to survey entire genomes even when we lack the computational resources necessary to load entire genomes into memory.



## Example data


vcfR comes with several example datasets.
Here we'll use them to create a small example genomic dataset.


```{r}
library(vcfR)
data(vcfR_example)
data("vcfR_test")
```


The vcfR_test dataset consists only of VCF data.
We'll fabricate a sequence and annotation for the sake of an example.


```{r}
library(ape)

dna.l <- vector('list', length=3)

dna.l[[1]] <- as.character(dna[1,1:length(dna)])
set.seed(123)
dna.l[[2]] <- sample( c('a','c','g','t'), size = getPOS(vcfR_test)[length(getPOS(vcfR_test))], replace = TRUE )
dna.l[[3]] <- sample( c('a','c','g','t'), size = getPOS(vcfR_test)[length(getPOS(vcfR_test))], replace = TRUE )
dna.l <- as.DNAbin(dna.l)

names(dna.l)[1] <- "Supercontig_1.50"
names(dna.l)[2] <- "Supercontig_1.5"
names(dna.l)[3] <- "Supercontig_1.10"
dna.l

gff2 <- gff[1:10,]
gff2[,1] <- "Supercontig_1.5"
gff3 <- gff[11:20,]
gff3[,1] <- "Supercontig_1.10"
```


I'm used to receiving reference and annotation data as a single file each.
We'll combine those data here for the example.


```{r}
gff <- rbind(gff, gff2, gff3)
rm(gff2)
rm(gff3)
```


I'm also used to having my VCF data seperated as one file per chromosome.
So we'll write our VCF data to file.


```{r}
write.vcf(vcf, file = "Supercontig_1.50.vcf.gz")
vcfR_test@fix[,'CHROM'] <- "Supercontig_1.5"
write.vcf(vcfR_test, file = "Supercontig_1.5.vcf.gz")
vcfR_test@fix[,'CHROM'] <- "Supercontig_1.10"
write.vcf(vcfR_test, file = "Supercontig_1.10.vcf.gz")
```


## Windowing through chromosomes


We now have a reference sequence and annotation information in memory and VCF data in files.
We are going to use vcfR to windowize each chromosome and write summaries for each window and each variant to file.
This is accomplished by first writing the files to disk and subsequently appending to each file.


```{r}
myFiles <- list.files(".", pattern = "Supercontig.*vcf.gz$")
myChroms <- unlist( lapply( strsplit(myFiles, "\\.vcf"), function(x){ x[1] } ) )

myWin.size <- 1e4
i <- 1
myChrom <- myChroms[i]

dna1 <- dna.l[myChrom]
gff1 <- gff[ gff[,1] == myChrom,]

vcf <- read.vcfR(myFiles[i], verbose = FALSE)
    
chrom <- create.chromR(name=myChrom, vcf=vcf, seq=dna1, ann=gff1, verbose = FALSE)
chrom <- proc.chromR(chrom, win.size = myWin.size, verbose=FALSE)

write.var.info(chrom, file = "pinf_ref_vars.csv", APPEND = FALSE)
write.win.info(chrom, file = "pinf_ref_wins.csv", APPEND = FALSE)

for(i in 2:length(myFiles)){
  myChrom <- myChroms[i]
  cat(myChrom)
  cat("\n")
  
  dna1 <- dna.l[myChrom]
  gff1 <- gff[ gff[,1] == myChrom,]

  vcf <- read.vcfR(myFiles[i], verbose = FALSE)

  chrom <- create.chromR(name=myChrom, vcf=vcf, seq=dna1, ann=gff1, verbose = FALSE)
  chrom <- proc.chromR(chrom, win.size = myWin.size, verbose=FALSE)

  write.var.info(chrom, file = "pinf_ref_vars.csv", APPEND = TRUE)
  write.win.info(chrom, file = "pinf_ref_wins.csv", APPEND = TRUE)

}

```


## Clean up

We wrote two files to disk for this example.
Typically these files would be data and would therefore be valuable.
In the context of this example, we no longer need them, so we'll delete them.
Make sure you do not delete your real data!


```{r, eval=FALSE}
unlink("Supercontig_1.5.vcf.gz")
unlink("Supercontig_1.10.vcf.gz")
unlink("Supercontig_1.50.vcf.gz")

unlink("pinf_ref_vars.csv")
unlink("pinf_ref_wins.csv")
```



